---
title: "A Gibbs Sampler"
author: "Christian Barz"
date: "created: 2021-06-16, reviewed: `r lubridate::today()`"
output: 
  github_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
```

# A bayesian regression

We consider the following regression

$$
\begin{align}
y &= 
\beta_0 + \beta_1 x + \epsilon\\
\epsilon &\sim 
N(0,1/\tau)
\end{align}
$$

To do a bayesian regression we need to set prior on the parameters. We do this 
by using conditionally conjugated priors. 
A *conditionally conjugated prior for a parameter $x$* for
a given likelihood gives a posteriori when conditioned on all parameters

$$
\begin{align}
\beta_i  & \sim N(\mu_i,1/\tau_i)\\
\tau & \sim Gamma(a,b)
\end{align}
$$

We assume that the hyperparameters $\mu_0,\mu_1,\tau_0,\tau_1,a,b$ are known 
constants.

By bayes rule the posterior distribution is proportional to

$$
p(\beta,\tau|y)\propto p(y|\beta,\tau) p(\beta) p(\tau)
$$

To do Gibbs sampling, we need the conditional posteriori distributions of the parameters.

$$
\beta_0|x,y,\beta_1,\tau \sim N(\frac{\tau_0\mu_0+\tau\sum_i(y_i-\beta_1x_i)}{\tau_0+n\tau},\frac{1}{\tau_0+n\tau})
$$
$$
\beta_1|x,y,\beta_0,\tau \sim 
N(\frac{
  \tau_1\mu_1 + 
  \tau\sum_i(y_i-\beta_0)x_i 
}{
  \tau_1+\tau\sum_ix_i^2
  }
,
\frac{1}{\tau_1+\tau\sum_ix_i^2}
)
$$

$$
\tau|\beta_0,\beta_1,a,b,x,y \sim
Gamma(a + \frac{N}{2}, b + \sum_i\frac{(y_i\beta_0-\beta_1x_i)^2}{2})
$$

## define conditional posterior probabilities

```{r}
sample_beta0 <- function(y,x,beta1,tau,tau0,mu0){
  N <- NROW(y)
  precision <- 1 / (tau0 + N * tau)
  mittelwert <- (tau0 * mu0 + tau * sum(y - beta1 * x) ) / precision
  
  return(rnorm(n = 1, 
               mean = mittelwert,
               sd = 1/ sqrt(precision))
         )
}
```


```{r}
sample_beta1 <- function(y,x,beta0,tau,tau1,mu1){
  N <- NROW(y)
  precision <- 1 / (tau1 + tau * sum(x * x))
  mittelwert <- (tau1 * mu1 + tau * sum((y-beta0) * x)) / precision
  
  return(rnorm(n = 1, 
               mean = mittelwert, 
               sd = 1/ sqrt(precision))
         )
}
```

```{r}
sample_tau <- function(y,x,beta0, beta1, a,b){
  N <- NROW(y)
  a_posteriori <- a + N/2
  b_posteriori <- b + sum((y - beta0 - beta1 * x)^2)/2
  return(
    rgamma(n = 1, 
           shape = a_posteriori,
           scale = b_posteriori)
  )
}
```



## gibbs sampler

```{r}
gibbs <- function(y,x, 
                  initialvalue, 
                  #conditional_posteriors,
                  hyperparameter,
                  iterations){
  
  
  samples <- array(data = NA, 
                   dim = c(iterations,
                           length(initialvalue)))
  
  colnames(samples) <- names(initialvalue)
  
  samples[1,] <- initialvalue %>%
    purrr::reduce(cbind)
  
  for(i in 2:iterations){
    
    samples[i,"beta0"] <- sample_beta0(y = y, 
                                       x = x, 
                                       beta1 = samples[i-1,"beta1"],
                                       tau = samples[i-1,"tau"],
                                       tau0 = hyperparameter[["tau0"]],
                                       mu0 =  hyperparameter[["mu0"]])
    
    samples[i,"beta1"] <- sample_beta1(y = y, x = x,
                           beta0 = samples[i,"beta0"], 
                           tau = samples[i-1,"tau"], 
                           tau1 = hyperparameter[["tau1"]],
                           mu1 = hyperparameter[["mu1"]]
                            )
    
    samples[i,"tau"] <- sample_tau(y = y, x = x,
                                   beta0 = samples[i,"beta0"], 
                                   beta1 = samples[i,"beta1"],
                                   a = hyperparameter[["a"]], 
                                   b = hyperparameter[["b"]]
                                   )
  }
  
  return(samples)
}
```


## test

### synthetic data

```{r synthetic data}
n <- 1
m <- 2
sigma <- 10.

# done a series of experiments and repeated it 4 times
x <- rep.int(x = seq(from = 0, to = 20, by = 1),
             times = 4)
y <- sapply(x, function(i){
  rnorm(n = 1,
        mean = m * i + n,
        sd = sigma)
})
```

### initial values, hyperparameters

```{r}
conditional_posteriors <- list(sample_beta0, 
                               sample_beta1,
                               sample_tau)
names(conditional_posteriors) <- c("beta0", "beta1", "tau")

initialvalue <- list(1,1,1)
names(initialvalue) <- c("beta0", "beta1", "tau")

hyperparameters <- list(0,1,0,1,2,2)
names(hyperparameters) <- c("mu0", "tau0",
                            "mu1", "tau1", 
                            "a", "b")
```



```{r}
gibbs(y = y, x = x, 
      initialvalue = initialvalue,
      hyperparameter = hyperparameters, 
      iterations = 10)
```

