---
title: "sbc"
author: "Christian Barz"
date: "created 2021-05-27, updated: `r lubridate::today()`"
output: 
  github_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
```

# to dos

- [ ] include a real sampler
- [ ] include a "real" model
- [ ] expand to a vector of random variables (auswertungsfunktionen der samples)

# sbc

## steuerungsparameter 

```{r}
L <- 100 # number of posterior draws
N <- 10000 # 
parameters <- 1 # additional parameters, e.g. predictor matrix or hyperparameters
```


## random variable to analyse the parameter

```{r}
# one-dimensional random variable (used in rank statistic)
f <- function(x){x}
```

## rank statistic

```{r}
# rank statistic , according to 4.1 in sbc talts
rank_statistic <- function(f, theta_L, theta){
  lapply(theta_L, function(x){
    f(x) < f(theta)
  }) %>%
    purrr::reduce(sum)
}
```

## prior (draws)

draws a sample from the prior. in general it has the form $P(\theta|x)$ as our prior may dependt on an additional variable, e.g. the predictors $x$.

This function has to be specified by the user and defines a test

```{r}
prior_draw <- function(parameters){
  theta <- rnorm(n=1, # draw one sample from a normal distribution
                 mean = 0, # with mean 0
                 sd = parameters) # and standard deviation 1
  return(theta)
}
```


## likelihood (draws)

in general $P(y|\theta,x)$
has to be specified by user and defines a test

```{r}
likelihood_draw <- function(theta, parameters){
  y <- rnorm(n = 1,
             mean = theta,
             sd = parameters)
  return(y)
}
```

## posterior (draws from the sampler)


we recall bayes theorem

$$
P(\theta|y,x) \propto P(y|\theta,x) \cdot P(\theta|x)
$$

a sampler now uses the right hand side to generate samples for the left 
hand side. this sentence motivates the parameter `theta` in the definition 
of the following function `sampler_draws`

in practice it will be replaced by the sampler we want to test.


```{r}
sampler_draws <- function(L,theta,y,parameters){
    theta_L <- rnorm(n=L, 
                     mean = y, 
                     sd = parameters)
}
```


## sbc

sbc consists in at least 2 steps

1. generate draws from the prior $\theta$ and $\theta_L$ and compute rank
statistics

2. analyse the output from step 1.

### generate values

```{r}
sbc <- function(prior_draw,
                likelihood_draw, 
                sampler_draws,
                f,
                N = 1000,
                L = 100,
                parameters
                ){
  
  sapply(1:N, function(i_){
    
    theta <- prior_draw(parameters)
    # compute rank statistic
    rank_statistic(f = f, 
                   theta_L = sampler_draws(L = L,
                                           theta = theta,
                                           y = likelihood_draw(theta, 
                                                               parameters), 
                                           parameters = parameters
                                           ), 
                   theta = theta)
  })
}
```

```{r}
values <- sbc(
  prior_draw = prior_draw, 
  likelihood_draw = likelihood_draw, 
  sampler_draws = sampler_draws,
  f = f,
  N = N,
  L = L,
  parameters = parameters
  )
```

### analyse values

```{r}
hist(values, breaks = L)
```


